---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Downloading source file. Windows RStudio requires mode="wb"
```{r}
sourceURL <- "http://d396qusza40orc.cloudfront.net/repdata/data/activity.zip"
localZipFile <- "activity.zip"
localFile <- "activity.csv"

download.file(sourceURL, dest=localZipFile, mode="wb")
## unpack
unzip(localZipFile,exdir=".")
```
Read data file into R, date is converted to POSIXlc type.
```{r}
## read
activity<-read.csv(file=localFile,header=TRUE, na.strings="NA",stringsAsFactors = FALSE)
## convert to date
activity$DT <- as.POSIXct(strptime(activity$date,format="%Y-%m-%d"),tz="UTC")
```

## What is mean total number of steps taken per day?
```{r warning=FALSE , message=FALSE}
library(dplyr)
```
Mean total number of steps per day requires aggregation by date. min(), max() sd() are not necessary.
```{r}
activity_grp1 <- group_by(activity,DT)

activity_mean1 <- activity_grp1 %>% 
                   summarise(steps_mean=mean(steps,na.rm=TRUE), steps_sd=sd(steps,na.rm=TRUE),
                             steps_median=median(steps,na.rm=TRUE),
                             steps_min=min(steps,na.rm=TRUE),steps_max=max(steps,na.rm.=TRUE))
```

```{r}
hist(activity_mean1$steps_mean,main="Mean total number of steps taken per day",xlab="mean total number of steps per day",col="blue")

hist(activity_mean1$steps_median,main="Median total number of steps taken per day",xlab="median total number of steps per day",col="red")
```
Mean and median per day
```{r}
# calculate y axis
yl<- c(  min( c(activity_mean1$steps_mean,activity_mean1$steps_median),na.rm=TRUE)  , max( c(activity_mean1$steps_mean,activity_mean1$steps_median),na.rm=TRUE ) )
## plot, but without x axis 
  plot(activity_mean1$DT,activity_mean1$steps_mean,type="l",col="blue",xlab="",ylab="", xaxt = "n", ylim=yl)
	par(new=T)
	plot(activity_mean1$DT,activity_mean1$steps_median,type="l",col="red",xlab="",ylab="", xaxt = "n", ylim=yl)
		## plot custom x axis
	axis.POSIXct(1,activity_mean1$DT, xlab="", format = "%Y-%m-%d")
	par(new=T)
	legend('topright','groups',c("Mean","median"), lty = c(1,1), col=c('blue','red'),bty="n")
	
```

## What is the average daily activity pattern?
To show daily activity pattern data will be aggregated by day for each 5min interval.
```{r warning=FALSE , message=FALSE}
library(dplyr)
```

```{r}
activity_grp2 <- group_by(activity,interval)

activity_mean2 <- activity_grp2 %>% 
                   summarise(steps_mean=mean(steps,na.rm=TRUE), steps_sd=sd(steps,na.rm=TRUE),
                             steps_min=min(steps,na.rm=TRUE),steps_max=max(steps,na.rm.=TRUE))
```

```{r}
plot(activity_mean2$interval, activity_mean2$steps_mean , type="l", xlab= "intervals", ylab= "mean number of steps", col="blue" , lwd=2)
```



## Imputing missing values
Source data contain many rows without value for steps column
```{r}
table(is.na(activity$steps))
```
NA values will be replaced by mean value of steps for that interval
```{r}
activity2<-activity
```
```{r}
for(i in 1:nrow(activity2)){
  if (is.na( activity2[i,"steps"]) ) {
    d_int<-activity2[i,"interval"]
    s<-activity_mean2[activity_mean2$interval==d_int,"steps_mean"]
    if ( !is.na(s)) {     
      activity2[i,"steps"]<-s  
    }
  }
}

```
```{r}
table(is.na(activity2$steps))
```
Mean total number of steps per day requires aggregation by date. min(), max() sd() are not necessary.
```{r warning=FALSE , message=FALSE}
library(dplyr)
```
```{r}
activity2_grp1 <- group_by(activity2,DT)

activity2_mean1 <- activity2_grp1 %>% 
                   summarise(steps_mean=mean(steps,na.rm=TRUE), steps_sd=sd(steps,na.rm=TRUE),
                             steps_median=median(steps,na.rm=TRUE),
                             steps_min=min(steps,na.rm=TRUE),steps_max=max(steps,na.rm.=TRUE))
```

```{r}
hist(activity2_mean1$steps_mean,main="Mean total number of steps taken per day",xlab="mean total number of steps per day",col="blue")

hist(activity2_mean1$steps_median,main="Median total number of steps taken per day",xlab="median total number of steps per day",col="red")
```
Mean and median per day
```{r}
# calculate y axis
yl<- c(  min( c(activity2_mean1$steps_mean,activity2_mean1$steps_median),na.rm=TRUE)  , max( c(activity2_mean1$steps_mean,activity2_mean1$steps_median),na.rm=TRUE ) )
## plot, but without x axis 
  plot(activity2_mean1$DT,activity2_mean1$steps_mean,type="l",col="blue",xlab="",ylab="", xaxt = "n", ylim=yl)
  par(new=T)
	plot(activity2_mean1$DT,activity2_mean1$steps_median,type="l",col="red",xlab="",ylab="", xaxt = "n", ylim=yl)
		## plot custom x axis
	axis.POSIXct(1,activity2_mean1$DT, xlab="", format = "%Y-%m-%d")
	par(new=T)
	legend('topright','groups',c("Mean","median"), lty = c(1,1), col=c('blue','red'),bty="n")
	
```
Peaks in median value are corresponding to days where all measures had steps equal NA.
```{r}
unique(activity[is.na(activity$steps),"DT"])
```

## Are there differences in activity patterns between weekdays and weekends?

```{r warning=FALSE , message=FALSE}
library(timeDate)
library(ggplot2)
```
```{r}
activity2$wkday <- ifelse(isWeekday(activity2$DT),"weekday","weekend")
activity2_grp2 <- group_by(activity2,interval,wkday)

activity2_mean2 <- activity2_grp2 %>% 
                   summarise(steps_mean=mean(steps), steps_sd=sd(steps),
                             steps_median=median(steps),
                             steps_min=min(steps),steps_max=max(steps))

```

```{r}
ggplot(data=activity2_mean2 , aes( interval , steps_mean) )  + geom_point() + facet_grid( . ~ wkday) 

```
